Demystifying the Computer Science PhD Admission in the US\\{\large A Guide for International Students

### 2023-12-05 demystify.tex
updates

The primary focus of the admissions committee is to \textbf{evaluate your background and interest in research} (a PhD in Computer Science is a research degree!), and determine if you would \textbf{fit into} the program. To evaluate your profile, we consider
the following key indicators, listed in order of importance.

@@

The most effective evidence of research ability is having \textbf{published papers in reputable international journals or conferences}.
Having published good papers is a sign that the applicant has successfully involved in research.

However, it is understandable that many students do not have the opportunities to publish in top places. Thus, general writings, even those under submissions or rejected, would still be good (and a lot better than nothing).  But be sure to upload your papers with your application and talk about them in your statement (see \S\ref{sec:research-statement}).  Note that local conferences and non-English journals or conferences do
not carry as much weight since their quality is often unknown to US faculty. However, if you have published in such places, you should still mention them in your statement and explain why they are good.

@@

\begin{commentbox}[Vu:]
  Many international students mention Scopus Q1, which consists of various journals from IEEE, Elsevier, and many other publishers.  I don't know/recognize many of journals listed in Scopus Q1. This might be something to be mindful of, as \textbf{CS} faculty might not be too familiar with Scopus or journals listed there, so devote sometime in your statement to discuss the significance of your papers.
\end{commentbox}

@@

Most CS PhD programs will require at least \textbf{two LORs}. Having a letter from an internationally recognized researcher can greatly strengthen your application. However, obtaining such letters
can be challenging for international students, who might not have much interactions with such experts. So it is fine to have a letter from people (e.g., profs, researchers, mentors) that know you well enough to talk about \emph{your specific research experience and capabilities}. This person can be a postdoc or even graduate student who has mentored you.

@@

Many students have letters written by themselves and signed by their professors. These have little
values. We can recognize them and will consider them weakness.
Also, many professors write generic letters for students, e.g., the students didn't do any
research or make any impression for the professor to write about. These letters are also not useful and considered weak.  It might be a good idea to directly ask if the prof. is willing to write a \emph{strong} letter for you. If not, then you should ask someone else.  For example, if a student I don't know well ask me to write letter for them, I will explicitly tell them I don't know them that well to write much about them, and such a short, generic, and weak letter will not help their profile.

@@

\paragraph{Always Waive your Right to Review the LORs}  Choosing not to look at a reference letter is pretty standard in job and school applications. When you waive your right to see the letter, it adds a layer of trust, showing you're confident in your choice of referees and that you're not trying to twist their words. It's also about keeping things open and honest between you and your letter writters, and encourages them to be real about your strengths and qualifications. Plus, it keeps things private.

@@

\paragraph{VEF 2.0} For Vietnamese students, it's worth mentioning about the \href{https://vef2.org/}{VEF2.0 program}, which has helped many good students in gaining acceptance to top PhD programs in the US. VEF2.0 follows an interesting model where US faculty members from leading institutions are invited to conduct rigorous interviews with VEF students and subsequently provide reference letters on their behalf. Despite the limited interaction between the interviewer and the interviewee (primarily confined to the interview itself), the impact of these reference letters can be positive as they have helped many students getting into top PhD programs in the US.

>>>>

The primary focus of the admissions committee is to \textbf{evaluate your background and interest in research} (a PhD in Computer Science is a research degree!), and determine if you would \textbf{fit into} the program. To evaluate your profile, we consider
the following key indicators.

@@

The most effective evidence of research ability is having \textbf{published papers in reputable international journals or conferences}.
Having papers published is a sign that the applicant has successfully involved in research.

\begin{commentbox}[Vu:]
    Many international students mention Scopus Q1, which consists of various journals from IEEE, Elsevier, and many other publishers.  I don't know/recognize many of journals listed in Scopus Q1. This might be something to be mindful of, as \textbf{CS} faculty might not be too familiar with Scopus or journals listed there, so devote some part in your statement to discuss the significance of your papers.
\end{commentbox}

However, it is understandable that many students do not have the opportunities to publish in top places. Thus, general writings, even those under submissions or rejected, would still be good (and much better than nothing).  But be sure to upload your papers with your application and talk about them in your statement (see \S\ref{sec:research-statement}).  Note that local conferences and non-English journals or conferences do
not carry as much weight since their quality is often unknown to US faculty. However, if you have published in such places, you should still mention them in your statement and explain why they are good.

@@

@@

@@

Most CS PhD programs will require at least \textbf{two LORs}. Having a letter from an internationally recognized researcher can greatly strengthen your application. However, obtaining such letters
can be challenging for international students, who might not have much interactions with such experts. So it is fine to have a letter from people (e.g., profs, researchers, mentors) that know you well enough to talk about \emph{your research experience and capabilities}. This person can be a postdoc who has mentored you.

\begin{commentbox}[Vu:]
    If you are from a good school from your country or your grading system is not standard. You can ask your reference writers to explain about that in their letters.  For example, "Bac Khoa" are the top universities in Vietnam for CS but few people outside Vietnam know about them.  So if you are from there, you should ask your reference writers to mention that in their letters.
\end{commentbox}

Many students have letters written by themselves and signed by their professors. Such letters have little
values and would be considered weak (you cannot even find someone cares enough about you to write a candid personal reference letter?).  Instead of the reference writer writing about you, in this case it is you who write about yourself (and they just sign the letter).

Also, many professors write generic letters for students, e.g., the students didn't do any
research or make any impression for the professor to write about. These letters are also not useful and considered weak.  It might be a good idea to directly ask if the prof. is willing to write a \emph{strong} letter for you. If not, then you should ask someone else.  For example, if a student I don't know well ask me to write letter for them, I will explicitly tell them I don't know them that well to write much about them, and such a short, generic, and weak letter will not help their profile.

@@

\paragraph{Always Waive your Right to Review the LORs}  Choosing not to look at a reference letter is pretty standard in school and job applications. When you waive your right to see the letter, it adds a layer of trust, showing you're confident in your choice of referees and that you're not trying to twist their words. It's also about keeping things open and honest between you and your letter writers, and encourages them to be real about your strengths and qualifications. Plus, it keeps things private.

@@

\paragraph{Reminding ref. writers to submit their letters} After you submit your application, you can tell your writers that you have submitted your application and remind them to submit their letters if they haven't done so. But don't send too many reminders as that can be annoying to the writers.

Note that most places only have deadlines for you, the applicant, but are very flexible with the letter writers (in many cases do not really give them any deadline or a very general one).  Also, many places do not begin the admission review process right after the deadline and instead won't start until the next semester (mid January).

\paragraph{VEF 2.0} For Vietnamese students, it's worth mentioning about the \href{https://vef2.org/}{VEF2.0 program}, which has helped many good students in gaining acceptance to top PhD programs in the US. VEF2.0 follows an interesting model where US faculty members from leading institutions are invited to conduct rigorous interviews with VEF students and subsequently provide reference letters on their behalf. Despite the limited interaction between the interviewer and the interviewee (primarily confined to the interview itself), these reference letters are generally helpful as they have helped many students getting into top PhD programs in the US.
